Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-08-31 Manuscript/01-Intro.Rmd
Update Real Data Analysis

There exists some proposals to generalized the classical GAM to accommodate high-dimensional applications. The regularized models that are primarily designed for linear models are used to fit GAMs by accounting for the group structure introduced when expanding smoothing functions.
<!-- These models assume sparsity among the predictors, i.e. most predictors have small or no effects on the outcome. Further development focuses on addressing the grouping structure among the predictors, where AMs and GAMs can be considered special cases. Various penalties are imposed on the basis function coefficients:  -->
Ravikumar et al. [@Ravikumar2009] extended the grouped lasso [@Yuan2006] to additive models; Huang et al. [@Huang2010] further developed adaptive grouped lasso for additive models; Wang et al. [@Wang2007] and Xue [@Xue2009] respectively applied grouped SCAD penalty [@Fan2001] to the additive models. Recent Bayesian hierarchical models are also used in the context of high-dimensional additive models. Various group spike-and-slab priors combining with computationally intensive Markov chain Monte Carlo (MCMC) algorithms [@Xu2015; @Yang2020] are proposed, where the application on AMs are by-products. Scheipl et al. [@Scheipl2012] proposed an ad hoc spike-and-slab structure prior, normal-mixture-of-inverse gamma, specifically for semi-parametric regression using MCMC. The normal-mixture-of-inverse gamma prior is mixture of t-distribution with an additional coefficients mixing vector, where the vector follows a mixture of two Gaussian distribution. Bai et al. [@Bai2020] was the first to apply group spike-and-slab lasso prior to AMs using a fast optimization algorithm, and further generalized the framework to GAMs[@Bai2021].

TODO: emphasize it doesn't accounting for smoothing penalty, and possibly lead to overshinkage, resulting wrong estimation of the smoothing function and prediction.

Regardless of the nice statistical properties, a practical and pivotal question, if non-linear effects exist, yet to be addressed by these methods. Besides Scheipl et al. [@Scheipl2012], most proposed methods adapt an 'all-in-all-out' strategy, i.e. either including or excluding the variable completely, rendering no space for within group selection. In addition, a common criticism received by these methods is that the basis function coefficients tend to be overly penalized due to the sparsity assumption directly applied on the basis function coefficients.[@Scheipl2013] It would be of special interest to develop a fast yet flexible generalized additive model.

To address these challenges, we propose a novel Bayesian hierarchical generalized additive model (BHAM) for high dimensional data analysis. Specifically, we impose a new bi-part spike-and-slab spline prior on the smooth functions to perform bilevel selection, where the linear and nonlinear spaces of smooth functions can be selected separately. The prior setup encourages a flexible solution, rending one of three possibilities: no effect, only linear effect, or nonlinear effect of a predictor. Additionally, two fast computing optimization-based algorithms are developed for high and ultrahigh-dimensional[@Fan2009] settings respectively. Meanwhile, to avoid the over-penalization of the basis function coefficients, smoothness penalties are incorporated in the model via re-parameterization. Smoothness penalties are commonly implemented in GAMs through smoothing and penalized regression splines. They are quadratic norms of the coefficients and allow locally adaptive penalties on each smoothing function. For the most common cases of cubic splines, the smoothness penalty conditioning on a smoothness penalty parameter $\lambda_j$ is a function of the integration of the second derivative of the spline function, expressed mathematically as
\begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{"}_j(x)dx = \lambda_j \bs \beta_j^T \bs S_j \bs \beta_j ,
\end{equation}
where $\bs S_j$ is a known smoothness penalty matrix, and $\bs \beta_j$ is the basis function coefficients. Smoothness penalties are also previously used in the spike-and-slab GAM [@Scheipl2012] and the sparsity-smoothness penalty [@Meier2009]. Moreover, incorporating the smoothness penalty allows the separation of the linear space of the smooth function, also know as the null space, from the nonlinear space. Marra and Wood [@Marra2011] introduced a re-parameterization of the basis function matrix based on the decomposition of the smoothness penalty matrix $\bs S$. The re-parameterization allows different penalties imposed on the null and nonlinear spaces separately, and further allows the sparsity in linear space and smoothness in nonlinear effect interpolation. In addition, the priors of the linear and nonlinear effects are independent, creating computational convenience and further improving the scalability of the model.

The proposed BHAM uses a spike-and-slab prior for function selections, but it substantially differs from previous spike-and-slab based GAMs, i.e. the spike-and-slab GAM [@Scheipl2012] and the SB GAM [@Bai2021] in many ways. First of all, the proposed spike-and-slab spline prior is a spike-and-slab lasso type prior using independent mixture double exponential distribution, compared to spike-and-slab GAM. Spike-and-slab lasso priors provide computational convenience during model fitting through optimization algorithms versus computational intensive sampling algorithms. They make fitting high- and ultrahigh-dimensional models more feasible without sacrificing performance in prediction and variable selection. Secondly, the optimization algorithms introduced in this article include EM-Coordinate Descent algorithm and EM-Iterative weighted least square algorithm for different utilities. The applications of the two aforementioned algorithm are fruitful in Bayesian hierarchical models in high-dimensional data analysis. [@Yi2012; @Rockova2014a; @Rockova2018] Even though SB-GAM with a group spike-and-slab lasso prior also uses EM-CD for model fitting, the proposed EM-IWLS provides uncertainty measures that could be further used for confidence bound calculation and hypothesis testing. EM-IWLS complements the lack of inference potential from EM-CD in a median and high dimensional setting. Furthermore, the proposed model addresses the incapability of bilevel selection in SB-GAM. Lastly, the R implementation of the Bayesian hierarchical additive model is available publicly via BHAM at [https://github.com/boyiguo1/BHAM](https://github.com/boyiguo1/BHAM), make transnational science more accessible.

@@

<!-- Lastly, The third aim of the dissertation work will focus on the software implementation of the proposed models and algorithms. We will implement the proposed models and algorithms in one of the most commonly used statistical computing environment R.  The R implementation of the Bayesian hierarchical additive model is available via a freely available R package, BHAM at (https://github.com/boyiguo1/BHAM)[https://github.com/boyiguo1/BHAM].  -->

>>>>

There exists some proposals to generalized the classical GAM to accommodate high-dimensional applications. The regularized models that are primarily designed for linear models are used to fit GAMs by accounting for the group structure introduced when expanding smoothing functions.
<!-- These models assume sparsity among the predictors, i.e. most predictors have small or no effects on the outcome. Further development focuses on addressing the grouping structure among the predictors, where AMs and GAMs can be considered special cases. Various penalties are imposed on the basis function coefficients:  -->
Ravikumar et al. [@Ravikumar2009] extended the grouped lasso [@Yuan2006] to additive models; Huang et al. [@Huang2010] further developed adaptive grouped lasso for additive models; Wang et al. [@Wang2007] and Xue [@Xue2009] respectively applied grouped SCAD penalty [@Fan2001] to the additive models. Recent Bayesian hierarchical models are also used in the context of high-dimensional additive models. Various group spike-and-slab priors combining with computationally intensive Markov chain Monte Carlo (MCMC) algorithms [@Xu2015; @Yang2020] are proposed, where the application on AMs are by-products. Bai et al. [@Bai2020] was the first to apply group spike-and-slab lasso prior to AMs using a fast optimization algorithm, and further generalized the framework to GAMs[@Bai2021]. As these methods focus on addressing the sparsity, it is very likely to overly penalize the basis function coefficients and produce inaccurate predictions and curve interpolation, particularly when complex signals are assumed and large number of knots are used. [@Scheipl2013] In addition, the proposed methods adapt an 'all-in-all-out' strategy, i.e. either including or excluding the variable completely, rendering no space for bi-level selection. Scheipl et al. [@Scheipl2012] proposed an ad hoc spike-and-slab structure prior, normal-mixture-of-inverse gamma prior, that address the previous challenges. But the model fitting relies on computational intensive MCMC algorithmic and creates scalability concern.
 <!-- specifically for semi-parametric regression using MCMC. The normal-mixture-of-inverse gamma prior is mixture of t-distribution with an additional coefficients mixing vector, where the vector follows a mixture of two Gaussian distribution. -->
It would be of special interest to develop a fast, flexible and accurate generalized additive model.

To address these challenges, we propose a novel Bayesian hierarchical generalized additive model (BHAM) for high dimensional data analysis. Specifically, we incorporate smoothness penalties in the model via re-parameterization of the smoothing function to avoid over-penalizing basis function coefficients. Smoothness penalties are commonly implemented in GAMs through smoothing and penalized regression splines. They are quadratic norms of the coefficients and allow locally adaptive penalties on each smoothing function. The smoothness penalty conditioning on a smoothness penalty parameter $\lambda_j$ is a function of the integration of the second derivative of the spline function, expressed mathematically as
\begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{\prime\prime}_j(x)dx = \lambda_j \bs \beta_j^T \bs S_j \bs \beta_j ,
\end{equation}
where $\bs S_j$ is a known smoothness penalty matrix, and $\bs \beta_j$ is the basis function coefficients. Smoothness penalties are also previously used in the spike-and-slab GAM [@Scheipl2012] and the sparsity-smoothness penalty [@Meier2009]. Moreover, incorporating the smoothness penalty allows the separation of the linear space of smooth functions, also know as the null space, from the nonlinear space. We then impose a new bi-part spike-and-slab spline prior on the smooth functions for bi-level selection such that the linear and nonlinear spaces of smooth functions can be selected separately. The prior setup encourages a flexible solution, rending one of three possibilities: no effect, only linear effect, or nonlinear effect of a predictor. In addition, two fast computing optimization-based algorithms are developed for high and ultrahigh-dimensional[@Fan2009] settings respectively.

<!-- Marra and Wood [@Marra2011] introduced a re-parameterization of the basis function matrix based on the decomposition of the smoothness penalty matrix $\bs S$. The re-parameterization allows different penalties imposed on the null and nonlinear spaces separately, and further allows the sparsity in linear space and smoothness in nonlinear effect interpolation. -->
 <!-- In addition, the priors of the linear and nonlinear effects are independent, creating computational convenience and further improving the scalability of the model. -->

The proposed BHAM uses a spike-and-slab prior for function selections, but it substantially differs from previous spike-and-slab based GAMs, i.e. the spike-and-slab GAM [@Scheipl2012] and the SB GAM [@Bai2021] in many ways. First of all, the proposed spike-and-slab spline prior is a spike-and-slab lasso type prior using independent mixture double exponential distribution, compared to spike-and-slab GAM. Spike-and-slab lasso priors provide computational convenience during model fitting through optimization algorithms versus computational intensive sampling algorithms. They make fitting high- and ultrahigh-dimensional models more feasible without sacrificing performance in prediction and variable selection. Secondly, the optimization algorithms introduced in this article include EM-Coordinate Descent algorithm and EM-Iterative weighted least square algorithm for different utilities. The applications of the two aforementioned algorithm are fruitful in Bayesian hierarchical models in high-dimensional data analysis. [@Yi2012; @Rockova2014a; @Rockova2018] Even though SB-GAM with a group spike-and-slab lasso prior also uses EM-CD for model fitting, the proposed EM-IWLS provides uncertainty measures that could be further used for confidence bound calculation and hypothesis testing. EM-IWLS complements the lack of inference potential from EM-CD in a median and high dimensional setting. Furthermore, the proposed model addresses the incapability of bi-level selection in SB-GAM. Lastly, the R implementation of the Bayesian hierarchical additive model is available publicly via BHAM at [https://github.com/boyiguo1/BHAM](https://github.com/boyiguo1/BHAM), make transnational science more accessible.

@@
